Boolean Operations for Surfaces in VTK Without External Libraries

### 2011-05-10 Document/LaTeX/Article.tex
Edited through section 2.

% The abstract should be a paragraph or two long, and describe the
% scope of the document.
\begin{abstract}
\noindent
A previous VTK Journal article provided a VTK class that enabled boolean operations on surfaces defined by \code{vtkPolyData} objects. While the capabilities introduced by the previous contribution satisfied numerous request on the vtk-developers and vtk-users email lists, the class could not be incorporated into VTK because of a licensing incompatibility with the GTS library used to perform the boolean operations. This article describes a set of classes than enable the same boolean operations as the previous contribution using only VTK classes. Our contribution also has some feature enhancements over the previous contribution, namely, preservation of mesh topology as well as passing point data and cell data through to outputs.

@@

Let $A$ and $B$ be sets. A third set $C$ can be defined from these sets via a \emph{boolean operation}. Three common boolean operations on sets are:

@@

In geometric modeling, a closed surface mesh $M_A$ may be considered the boundary of a space-filling set of 3D points $A$. Then given a second surface mesh $M_B$, a boolean operation may be applied to obtain a third surface mesh $M_C$. These simple operations can be used to define complex geometries.

Rather than convert two mesh representations to a set of points, perform the boolean operations on those points, and reconstruct a third mesh from the resulting points, direct operations on the bounding geometry based on signed distance fields and mesh-to-mesh clipping can be used. This article describes an implementation of those methods in VTK.

@@

The distance field $f(\x)$ of a mesh is defined as the distance from $\x$ to the nearest point on the mesh. The signed distance field is similar to the distance field, but the sign of the field is negative for locations $\x$ inside the mesh and positive for locations $\x$ outside the mesh. To determine whether $\x$ is inside or outside the mesh, a vector $\v = \x - \m$ is defined where $\m$ is the nearest point on the mesh. B\ae rentzen and Aan\ae s proved that the sign of the dot product of $\v$ with the \emph{angle-weighted pseudo-normal} of the mesh at $\m$ is the same as the sign of the distance field for closed, orientable 2-manifold surfaces in 3d Euclidean space \cite{Baerentzen2005}.

The angle-weighted pseudo-normal is defined separately for faces, edges, and points in a mesh. For faces, it is simply the normal of the triangle. For an edge, it is the average of the faces that share the edge. For a point, it is the average of the angle-weighted normals from each face defined by the point where the weight is the angle between the two face edges incident to the point.

We introduce the \code{vtkImplicitPolyData} class that defines an implicit signed distance function given an input \code{vtkPolyData}. This class is a subclass of \code{vtkImplicitFunction}, so it can be used by classes that operate on \code{vtkImplicitFunction} objects (see, e.g., \code{vtkSampleFunction}). The function value is the signed distance, and its gradient value is the angle-weighted pseudo-normal.

To support boolean operations, we introduce another class, \code{vtkPolyDataDistance}, that computes the distance from the first input \code{vtkPolyData} to a second. Optionally, the distance field from the second input \code{vtkPolyData} to the first can also be computed. These distances are computed at point locations and are stored as point data.

@@

\begin{enumerate}
\item The three points that define the cell.
\item The boundary lines of the cell.
\item The split intersection lines that lie in the cell.
\item The points that define the lines in 3.
\item Points from neighboring cells that lie on a cell edge but which were not already added in 4.
\end{enumerate}

The lines collected above are used to constrain the new triangulation of the points. That is to say, these lines should be present in the output mesh. Special processing of the boundary lines is required to achieve the desired split. Specifically, boundary lines must be split at points that lie on them.

The \code{vtkDelaunay2D} class can be used to achieve a constrained triangulation. A transformation of the points collected above that rotates them to the XY-plane is calculated is computed and set as the transform for the \code{vtkDelaunay2D} object. A new \code{vtkPolyData} object that contains the collected points and lines is defined and passed as both the input and constraint source to a \code{vtkDelaunay2D} object. The line point indices are renumbered to point to their locations in the new \code{vtkPolyData} object, and a map back to their original indices is created to remap the output cell indices from the triangulation to the point indices in the output mesh.

@@

You need to have the following software installed:

% The {itemize} environment uses a bullet for each \item.  If you want the 
% \item's numbered, use the {enumerate} environment instead.
\begin{itemize}
  \item  Insight Toolkit 2.4.
  \item  CMake 2.2
\end{itemize}

Note that other versions of the Insight Toolkit are also available in the
testing framework of the Insight Journal. Please refere to the following page
for details

\url{http://www.insightsoftwareconsortium.org/wiki/index.php/IJ-Testing-Environment}

% The preceding sections will have been written in a gentler,
% introductory style.  You may also wish to include a reference
% section, documenting all the functions/exceptions/constants.
% Often, these will be placed in separate files and input like this:

\appendix

\section{This is an Appendix}

To create an appendix in a Insight HOWTO document, use markup like
this:

\begin{verbatim}
\appendix

\section{This is an Appendix}

To create an appendix in a Insight HOWTO document, ....

\section{This is another}

Just add another \section{}, but don't say \appendix again.
\end{verbatim}

>>>>

% The abstract should be a paragraph or two long, and describe the
% scope of the document.
\begin{abstract}
\noindent
We have written a set of classes than enable boolean operations on surface meshes defined in \code{vtkPolyData} objects using only VTK classes. Our contribution offers some feature enhancements over previous solutions, namely, preservation of mesh topology and passing point data and cell data through to outputs.

@@

Let $A$ and $B$ be sets. A \emph{boolean operation} can be used to define a third set $C$. Three common boolean operations on sets are:

@@

In geometric modeling, a surface mesh $M_A$ may be considered the boundary of a space-filling set of 3D points $A$ if $M_A$ is a closed, orientable 2-manifold surface in 3D. Given a second surface mesh $M_B$ with the same properties, a boolean operation may be applied to obtain a third surface mesh $M_C$. These simple operations can be used to define complex geometries.

Rather than convert two mesh representations to a set of points, perform the boolean operations on those points, and reconstruct a third mesh from the resulting points, direct operations on the bounding geometry based on signed distance fields and mesh-to-mesh clipping can be used. This article describes an implementation of those methods to realize boolean operations on surfaces in VTK.

@@

The distance field $f(\x)$ of a mesh is defined as the distance from $\x$ to the nearest point on the mesh. The signed distance field is similar to the distance field, but the sign of the field is negative for locations $\x$ inside the mesh and positive for locations $\x$ outside the mesh. To determine whether $\x$ is inside or outside the mesh, a vector $\v = \x - \m$ is defined where $\m$ is the nearest point on the mesh. The sign of the distance field is the same as the dot product of $\v$ with the \emph{angle-weighted pseudo-normal} of the mesh at $\m$ when the mesh is a closed, orientable 2-manifold surfaces in 3D Euclidean space \cite{Baerentzen2005}.

The angle-weighted pseudo-normal is defined separately for faces, edges, and points in a mesh. For faces, it is simply the normal of the triangle. For an edge, it is the average of the normals for the faces that share the edge. For a point, it is the sum of the angle-weighted normals from each face where the weight for a face normal is the angle between the two edges of that face incident to the point.

We introduce the \code{vtkImplicitPolyData} class that defines a signed distance function given an input \code{vtkPolyData}. This class is a subclass of \code{vtkImplicitFunction}, so it can be used by classes that operate on \code{vtkImplicitFunction} objects (see, e.g., \code{vtkSampleFunction}). The method XXXX is overridden to return the signed distance, and XXXX is overridden to return the angle-weighted pseudo-normal.

To support boolean operations, we introduce another class, \code{vtkPolyDataDistance}, that computes the distance from the first input \code{vtkPolyData} to a second. Optionally, the distance field from the second input \code{vtkPolyData} to the first can also be computed. These distances are computed at points in the mesh and are stored as a point data field named ``Distance''.

@@

\begin{enumerate}
\item The three points that define the cell.
\item The boundary lines of the cell.
\item The split intersection lines that lie in the cell (if any).
\item The points that define the lines in 3.
\item Points from neighboring cells that lie on a cell edge but which were not already added in 4.
\end{enumerate}

The lines collected above are used to constrain the new triangulation of the points. That is to say, these lines should be present in the output mesh. Special processing of the boundary lines is required to achieve the desired split. Specifically, boundary lines must be split at points that lie on them. Boundary line splitting is achieved by sorting points on the boundary lines according to the angle between the first edge of the cell and the vector formed by subtracting the cell center from the point and connecting points adjacent to each other with lines. Because the last boundary point is connected to the first, the absolute order of the boundary points is not important.

The \code{vtkDelaunay2D} class can be used to determine a triangulation from the cell, surface intersection lines, and boundary points and lines. A transformation of the points collected above that rotates them to the XY-plane is calculated is computed and set as the transform for the \code{vtkDelaunay2D} object. A new \code{vtkPolyData} object that contains the collected points and lines is defined and passed as both the input and constraint source to a \code{vtkDelaunay2D} object. The line point indices are renumbered to point to their locations in the new \code{vtkPolyData} object, and a map back to their original indices is created to remap the output cell indices from the triangulation to the point indices in the output mesh. If a point in the surface intersection is within some small tolerance distance from a point in the surface mesh, the point in the surface mesh is used instead.

@@

The source code contributed with this article has been build against VTK 5.8.
